# Week 7

## `with` and `table`

The data always have 0 and 1 indicating exposure/disease/confounder status, to get contigency tables for further calculation, we use `with` and `table`.

```{r}
dat <- foreign::read.dta("./_data/diet.dta")
head(dat)
```

### 2 way table

- `table` uses the cross-classifying factors to build a contingency table of the counts at each combination of factor levels.

  - The input data for `table` should be one or more objects which can be interpreted as factors.
  
  - To create a two by two table for exposure and disease: `table(data$exposure, data$disease)`

```{r}
table(dat$diet, dat$mort)
```

- `with(data, expr, ...)`: evaluate an R expression in an environment constructed from data

  - `data`: data to use for constructing an environment. 
  - `expr`: expression to evaluate, in this case we use `table()`

```{r}
with(dat, table(diet,mort))
```

- The same results as using `table()` function only

  - As the environment has been constructed, there's no need to use `dat$` to tell R how to find the factor.
  

### 3 way table

- Similar method as 2 way table, but indicate the third factor using `by = `

  - To create a three way table for exposure and disease: `with(data, table(exposure, disease, by = confounder))`

```{r}
table3 <- with(dat, table(diet, mort, by = act))
table3
```


## `mantelhaen.test`

> Performs a **Cochran-Mantel-Haenszel chi-squared test** of the null that two nominal variables are conditionally independent in each stratum, assuming that there is no three-way interaction.

- To test the null hypothesis that the exposure is independent of the disease when adjusting  for confounding, we can use  

```{}
mantelhaen.test(x, y = NULL, z = NULL,
                alternative = c("two.sided", "less", "greater"),
                correct = TRUE, exact = FALSE, conf.level = 0.95)
```

- Input data: 

  - A 3-dimensional contingency table in array form as `x`
  - Or three factor objects with at least 2 levels as `x`, `y`, and `z`.
  
- `correct = TRUE`: Whether to apply continuity correction when computing the test statistic.

- `exact = FALSE`: Whether the Mantel-Haenszel test or the exact conditional test (given the strata margins) should be computed.

```{r}
# example
mantelhaen.test(table3)
```

