--- 
title: "Coding Guide"
subtitle: "PH 241: Statistical Analysis of Categorical Data"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: ""
---

```{r include=FALSE}
library(dplyr)
library(epiR)
library(epitools)
library(foreign)
```

# Packages

- we always include a setup chunk to include packages we use in the rmarkdown file, these are packages we usually use in this course:

1. dplyr[^1]

> dplyr is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges.

```{r eval=FALSE}
library(dplyr)
mutate()    # adds new variables that are functions of existing variables
select()    # picks variables based on their names.
filter()    # picks cases based on their values.
summarise() # reduces multiple values down to a single summary.
arrange()   # changes the ordering of the rows.
```



2. epiR[^2]

> Tools for the analysis of epidemiological and surveillance data. 

> Contains functions for directly and indirectly adjusting measures of disease frequency, quantifying measures of association on the basis of single or multiple strata of count data presented in a contingency table, computation of confidence intervals around incidence risk and incidence rate estimates and sample size calculations for cross-sectional, case-control and cohort studies. 

```{r eval=FALSE}
library(epiR)
epi.2by2() # Computes summary measures of risk and a chi-squared test for difference in a 2 by 2 table
```

3. epitools[^3]

> Epidemiology Tools

> Tools for training and practicing epidemiologists including methods for two-way and multi-way contingency tables.

```{r eval=FALSE}
library(epitools)
epitable() # Create r x c contigency table for r exposure levels and c outcome levels
epitab() # Calculates risks, RR, OR, and CIs for epidemiologic data
```

4. foreign[^4]

> Reading and writing data stored by some versions of 'Epi Info', 'Minitab', 'S', 'SAS', 'SPSS', 'Stata', 'Systat', 'Weka',
and for reading and writing some 'dBase' files.

```{r eval=FALSE}
library(foreign)
read.dta() # Reads a file in Stata version 5–12 binary format into a data frame
```



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

[^1]: Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: A Grammar of Data Manipulation. R package version 1.0.2. https://CRAN.R-project.org/package=dplyr

[^2]: Mark Stevenson, Evan Sergeant with contributions from Telmo Nunes,
  Cord Heuer, Jonathon Marshall, Javier Sanchez, Ron Thornton, Jeno
  Reiczigel, Jim Robison-Cox, Paola Sebastiani, Peter Solymos,
  Kazuki Yoshida, Geoff Jones, Sarah Pirikahu, Simon Firestone, Ryan
  Kyle, Johann Popp, Mathew Jay and Charles Reynard. (2020). epiR:
  Tools for the Analysis of Epidemiological Data. R package version
  2.0.17. https://CRAN.R-project.org/package=epiR

[^3]: Tomas J. Aragon (2020). epitools: Epidemiology Tools. R package
  version 0.5-10.1. https://CRAN.R-project.org/package=epitools
  
[^4]: R Core Team (2020). foreign: Read Data Stored by 'Minitab', 'S',
  'SAS', 'SPSS', 'Stata', 'Systat', 'Weka', 'dBase', .... R package
  version 0.8-81. https://CRAN.R-project.org/package=foreign
