---
title: "Week 4 Coding Guide"
subtitle: PH241
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(epiR)
library(dplyr)
library(kableExtra)
library(foreign)
library(here)
```

### epi.2by2

- Summary measures for count data presented in a 2 by 2 table

```{r eval=FALSE}
library(epiR)
epi.2by2(dat, 
         method = c("cohort.count", "cohort.time", "case.control", "cross.sectional"), 
         conf.level = 0.95, 
         units = 100, 
         outcome = c("as.columns","as.rows"))
```

- `dat`: an object of class `table` containing the individual cell frequencies.

- `method`: a character string indicating the study design on which the tabular data has been based.

  - Based on the study design specified by the user, appropriate measures of association, measures of effect in the exposed and measures of effect in the population are returned by the function.

```{r echo=FALSE}
methods <- data.frame("cohort.count" = c(rep("$\\checkmark$",6), rep("$\\checkmark$",5)), 
                      "case.control" = c("",rep("$\\checkmark$",2),"AFest","$\\checkmark$", "PAFest",rep("$\\checkmark$",3),"","$\\checkmark$"),
                      row.names = c("RR", "OR", "ARisk", "AFRisk", "PARisk",  "PAFRisk", 
                                    "chisq.strata", "chisq.crude", "chisq.mh",
                                    "RR.homog", "OR.homog")
  
)

kbl(methods) %>%
  #kable_paper("striped", full_width = F) %>%
  kable_classic() %>%
  pack_rows("Measures of association", 1, 2) %>%
  pack_rows("Measures of effect in the exposed", 3, 4) %>%
  pack_rows("Measures of effect in the population", 5, 6) %>%
  pack_rows("chi-squared test", 7, 9) %>%
  pack_rows("Mantel-Haenszel (Woolf) test of homogeneity", 10, 11)
```


- `conf.level`: magnitude of the returned confidence intervals. The default value is 0.95.

- `units`: multiplier for prevalence and incidence (risk or rate) estimates. Default value is 100.

  - `unit=1` Outcomes per population unit 
  
  - `unit=100` Outcomes per 100 population unit 
  
  - The multiplier is applied to Attributable risk, Attributable risk in population, etc.

- `outcome`: how the outcome variable is represented in the contingency table.

### Example

```{r}
# Example in A01
birthwt <- data.frame("Low" = c(21054, 27126), "Normal" = c(14442, 3804294), row.names = c("Dead at Year 1", "Alive at Year 1"))
birthwt
```

```{r}
epi.2by2(birthwt, 
         method = "cohort.count", 
         conf.level = 0.95, 
         units = 100, 
         outcome = "as.rows") # the outcome is represented as rows
```

### chisq.test

- performs chi-squared contingency table tests and goodness-of-fit tests.

```{r eval=FALSE}
chisq.test(x, y = NULL, correct = TRUE,
           p = rep(1/length(x), length(x)), rescale.p = FALSE,
           simulate.p.value = FALSE, B = 2000)
```

- `x, y = NULL`: the input can be two numeric vectors as x and y (can both be factors), or a matrix as x.

- `correct`: a logical indicating whether to apply **continuity correction**.

#### output components

```{r}
chi <- chisq.test(birthwt)
chi
# use dollar sign($) to specify output
chi$statistic # the value the chi-squared test statistic
chi$parameter # the degrees of freedom
chi$p.value
chi$method
chi$data.name
chi$observed
chi$expected  # the expected counts under the null hypothesis
chi$residuals # the Pearson residuals, (observed - expected) / sqrt(expected).
chi$stdres    # standardized residuals, (observed - expected) / sqrt(V)
```



### pchisq

- Density, distribution function, quantile function and random generation for the chi-squared ($\chi^2$) distribution with `df` degrees of freedom and optional non-centrality parameter `ncp`.

```{r eval=FALSE}
pchisq(q, df, ncp = 0, lower.tail = TRUE, log.p = FALSE)
```

- `q`: (vector of) quantile(s).

- `df`: degrees of freedom (non-negative, but can be non-integer).

- `lower.tail`: logical; if TRUE (default), probabilities are $P[X \le x]$, otherwise, $P[X > x]$.

#### Example

To find the p-value that corresponds with a $\chi^2$ test statistic of 7 from a test with one degree of freedom:

```{r}
pchisq(7, df=1, lower.tail = FALSE)
```

```{r echo=FALSE}
x <- rchisq(100000, df = 1)
hist(x, 
     breaks = 'Scott', 
     freq = FALSE, 
     xlim = c(0,10), 
     ylim = c(0,1),
     xlab = '', 
     main = 'Histogram for chi-squared distributions with 1 degree of freedom (df)', cex.main=0.9)

curve(dchisq(x, df = 1), from = 0, to = 15, n = 5000, col= 'orange', lwd=2, add = T)

abline(v = 7, col="red")
```

- The P-value is the probability of observing a sample statistic as **extreme** as the test statistic -- the area to the **left** of the red line -- "upper" tail

- We always set `lower.tail = FALSE` when calculating P-value of $\chi^2$ test.

- Another method to get the p-value

```{r}
# default: lower.tail = TURE
1 - pchisq(7, df=1)
```




